---
title: "Homework 2"
author: "Wayne Monical wem2121"
date: "2024-10-01"
output: github_document
---


```{r, results = 'hide'}
library(tidyverse)
```




## Problem 1

We begin by importing and cleaning the subway data set

```{r, results= 'hide'}
subway = read_csv(
  "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |> 
  janitor::clean_names() |> 
  select(line, station_name, station_la)
```


```{r}
subway_cols = c("line", "station_name", "station_latitude", "station_longitude",
                   "route1", "route2", "route3", "route4", "route5", "route6", "route7", 
                   "route8", "route9", "route10", "route11", "entrance_type", "entry",
                   "exit_only", "vending", "ada",    "ada_notes", "entrance_latitude",
                   "entrance_longitude")

subway = subway |> select(all_of(subway_cols))
```

We see here that `entry`, `exit_only`, and `vending` are character vectors, and `ada` is a logical vector
```{r}
logical_vars = c('entry', 'exit_only', 'vending', 'ada')

subway |> 
  select(all_of(logical_vars)) |> 
  summary()
```

Inspecting the unique values of each character, we can find the character strings that correspond to each logical value.
```{r}
char_vars = c('entry', 'exit_only', 'vending')

for(var in char_vars){
  print(var)
  
  subway |> 
    pull(var) |> 
    unique() |> 
    print()
}
```

Using the `mutate()` function, we can reassign these variables to be logical based on the character strings that we found.
```{r}
subway = subway |> 
  mutate(
    entry = entry == 'YES',
    exit_only = !is.na(exit_only),
    vending = vending == 'YES')
```


```{r}

```


```{r}
head(subway)
```


```{r}
colnames(subway)
```


```{r}
head(subway)
```


```{r}
subway |> pull('station_name') |> unique() |> length()
```


```{r}

```

