---
title: "Homework 2"
author: "Wayne Monical wem2121"
date: "2024-10-01"
output: github_document
---


```{r, results = 'hide'}
library(tidyverse)
library(readxl)
```




## Problem 1

### Data Cleaning

We begin by importing and cleaning the subway data set. This data set contains information on the organization, location, and amenities in the New York city subway system. The data is tidy. The data is consistent, i.e. each row of the data frame contains information on one subway entrance or exit. It is structured, i.e. every variable, such as station, line, and latitude, has its own column. Finally, all values have their own cell, i.e. there is no value concatenation. 

```{r}
subway = read_csv(
  "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |> 
  janitor::clean_names()
```


```{r}
subway_cols = c("line", "station_name", "station_latitude", "station_longitude",
                   "route1", "route2", "route3", "route4", "route5", "route6", "route7", 
                   "route8", "route9", "route10", "route11", "entrance_type", "entry",
                   "exit_only", "vending", "ada",    "ada_notes", "entrance_latitude",
                   "entrance_longitude")

subway = subway |> select(all_of(subway_cols))
```

We see here that `entry`, `exit_only`, and `vending` are character vectors, and `ada` is a logical vector
```{r}
logical_vars = c('entry', 'exit_only', 'vending', 'ada')

subway |> 
  select(all_of(logical_vars)) |> 
  summary()
```

Inspecting the unique values of each character, we can find the character strings that correspond to each logical value.
```{r}
char_vars = c('entry', 'exit_only', 'vending')

for(var in char_vars){
  print(var)
  
  subway |> 
    pull(var) |> 
    unique() |> 
    print()
}
```

Using the `mutate()` function, we can reassign these variables to be logical based on the character strings that we found.
```{r}
subway = subway |> 
  mutate(
    entry = entry == 'YES',
    exit_only = !is.na(exit_only),
    vending = vending == 'YES')
```


There are 465 distinct subway stations.
```{r}
subway |> select('station_name', 'line') |> distinct() |> nrow()
```

We must also standardize the route variables as characters. 
```{r}
subway = subway |> 
  mutate(
    route1 = as.character(route1),
    route2 = as.character(route2),
    route3 = as.character(route3),
    route4 = as.character(route4),
    route5 = as.character(route5),
    route6 = as.character(route6),
    route7 = as.character(route7),
    route8 = as.character(route8),
    route9 = as.character(route9),
    route10 = as.character(route10),
    route11 = as.character(route11),
  )
```



### Q & A

There are 84 ADA compliant subway stations. 
```{r}
subway |> filter(ada) |> select('station_name', 'line') |> distinct() |> nrow()
```

73% of station entrances and exits without vending allow entrance. 
```{r}
subway |> filter(!vending) |> pull(exit_only) |> summary()

133 / (50 + 133)
```

### Pivoting 
 
 Here, we pivot the route variables to a longer format and drop the rows that do not have a route. 
```{r}
subway= subway |> 
  pivot_longer(
    route1:route11,
    names_to = 'route_number',
    values_to = 'route'
  ) |> 
  filter(!is.na(route))
```

Using the code below, we find that 17 of the 43 of the stations on the A, 39%, are ADA compliant.  
```{r}
subway |> 
  filter(route == 'A') |> 
  select('line', 'station_name', 'ada') |>
  distinct() |> 
  pull(ada) |>
  summary()

107 / (166 + 107)
```

## Problem 2

Reading in the 2024 Mr Trash Wheel sheet. 
```{r}
trash = 
  read_excel(
    path = 'data/202409 Trash Wheel Collection Data.xlsx',
    sheet = 'Mr. Trash Wheel',
    range = 'A2:N653') |> 
  janitor::clean_names()
```



```{r}

```


